One-arm spirals in self-gravitating dead zones in protoplanetary discs

### 2014-10-21 paper/model.tex
tex edits, first figures

\subsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which circumvents time-step
contraint set by the inner disc boundary. 
The 2D Poisson
equation is solved in integral form using Fast Fouier Transform. This
requires a softening length $\epsilon_g H$, where $\epsilon_g$ is an
input parameter.

@@

\section{Fiducial simulations}
Here we describe one numerical simulation from each code to give an
overall picture of disc evolution. In the next section, we explore the
effect of disc parameters using 2D simulations.

The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D
generalisations. The hydrodynamic variables are written as 
\begin{align}
  f(R,\phi) &= \sum_{m=-\infty}^{\infty}f_m(R)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$. The
mode amplitude is 
\begin{align}
  A_m = \frac{|C_m|}{M_{d0}}
\end{align}
where $M_{d0}$ is the initial disc mass and 
\begin{align}
  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
\end{align}
with integration limits chosen later.

The total disc angular momentum in $R\in[R_1,R_2]$ is
\begin{align}
  J &= \int_{R_1}^{R_2}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_1}^{R_2}R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_1}^{R_2} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR, \notag\\
  &= \sum_{m=0}^\infty 2\pi \int_{R_1}^{R_2} j_m R dR = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the angular momentum 
associated with the $m^\mathrm{th}$ mode, and $j_m$ its density.

\subsection{WKB perturbations}
We will use results from linear density wave theory to set up
perturbations. 
Here, we follow conventions in \cite{shu91}. Recall the
dispersion relation for tightly-wound density waves of the form
$\exp{(\ii \omega t - m \phi + kR)}$ in a razor-thin
isothermal disc, 
\begin{align}\label{dispersion}
  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $\omega$ is the mode frequency, $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, and
$k$ is a real wavenumber.

If we specify the pattern speed $\Omega_p \equiv \omega/m$,
Eq. \ref{dispersion} can be solved for $|k|$,
\begin{align}\label{wavenumber}
  \frac{|k|}{k_T} = \frac{2}{Q^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
\begin{align}
  k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
\end{align}
is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
\nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. 
We use Eq. \ref{wavenumber} to set trailing long waves as
perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
    Q^2(1-\nu^2)}\right]$.  
We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.

The amplitude of the surface density and velocity perturbations are
related by
\begin{align}
  &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
  &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
  v_\phi=0. 
\end{align}
We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
v_\phi)$ accordingly.

%Having chosen the wave frequency and set the wavenumber according to
%the above,

>>>>

\subsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which circumvents time-step
contraint set by the inner disc boundary. 
The 2D Poisson equation is solved in integral form 
\begin{align}\label{2d_grav}
  \Phi_{d,z=0}(R,\phi) = - \int
  \frac{G\Sigma(R^\prime,\phi^\prime)}{\sqrt{R^2+R^{\prime 2} -
      2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}} 
\end{align}
using Fast Fouier Transform (FFT), where $\epsilon_g$ is a softening
length to prevent a numerical singularity and the integration is taken
over the disc. The FFT approach requires
$\epsilon_g\propto R$ \citep{baruteau08}.

@@

\section{Diagnostics}

\subsection{Evolution of non-axisymmetirc modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi) &= \sum_{m=-\infty}^{\infty}f_m(R)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

The non-axisymmetric surface density can be visualised with
\begin{align}
  \Delta \Sigma = \frac{\Sigma - \Sigma_{00}}{\Sigma_{00}}, 
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$, while an individual mode as
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right].
\end{align}

The global mode amplitude is 
\begin{align}
  A_m = \frac{|C_m|}{M_{d0}}
\end{align}
where $M_{d0}$ is the initial disc mass and 
\begin{align}
  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
\end{align}
with integration limits chosen later.

The time evolution of the $m^\mathrm{th}$ mode can be characterized by
assuming 
$\Sigma_m\propto\exp{(\ii \sigma t)}$, where
$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
the real frequency and $\gamma$ is the growth rate.
We estimate the growth rate as
\begin{align}
  \gamma = \frac{d\ln{A_m}}{dt},
\end{align}
and we estimate the real frequency by calculating the pattern speed
$\Omega_p = \omega/m$,
\begin{align}
  \Omega_p = \frac{d\phi_\mathrm{max}}{dt},
\end{align}
where $\phi_\mathrm{max}$ is the azimuthal co-ordinate of
$\mathrm{max}(\Delta\Sigma_m)$. 
%\begin{align}
%  \mathrm{max}\left\{2 \real\left[ \Sigma_m \exp{(\ii m
%        \phi)}\right]\right\},

%other methods possible too

\subsection{Angular momentum decomposition}
The total disc angular momentum in $R\in[R_1,R_2]$ is
\begin{align}
  J &= \int_{R_1}^{R_2}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_1}^{R_2}R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_1}^{R_2} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR, \notag\\
  &= \sum_{m=0}^\infty \int_{R_1}^{R_2}  j_m(R) dR = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the angular momentum 
associated with the $m^\mathrm{th}$ mode, and $j_m$ the associated
angular momentum per unit length.

\subsection{Linear density waves}
We will also use linear density wave theory to interpret
simulation data. Here, we follow conventions in \cite{shu91} and quote
results therein. The dispersion relation for tightly-wound density
waves of the form 
$\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
isothermal disc is 
\begin{align}\label{dispersion}
  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $\kappa^2 = R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, 
and $k$ is a real wavenumber such that $|kR|\gg1$.

Given the pattern speed $\Omega_p$,
Eq. \ref{dispersion} can be solved for $|k|$, 
\begin{align}\label{wavenumber}
  \frac{|k|}{k_T} = \frac{2}{Q^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
\begin{align}
  k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
\end{align}
is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency.  The upper (lower) signs in
Eq. \ref{wavenumber} correspond to short (long) waves.

The \emph{co-rotation radius} $R_c$ is where the pattern speed matches
the fluid rotation,
\begin{align}
  \Omega(R_c) = \Omega_p.
\end{align}
\emph{Lindblad resonances} $R_L$ occurs where
\begin{align}
  \nu^2(R_L) = 1. 
\end{align}
Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
\begin{align}
  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
\end{align}
According to Eq. \ref{wavenumber}, purely wave-like solutions with
real $k$ are only possible when $Q^2(1-\nu^2)\leq1$.

% We use Eq. \ref{wavenumber} to set trailing long waves as
% perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
%     Q^2(1-\nu^2)}\right]$.  
% We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.

% The amplitude of the surface density and velocity perturbations are
% related by
% \begin{align}
%   &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
%   &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
%   v_\phi=0. 
% \end{align}
% We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
% v_\phi)$ accordingly.

% %Having chosen the wave frequency and set the wavenumber according to
% %the above,

\section{Fiducial simulations}
Here we describe one numerical simulation from each code to give an
overall picture of disc evolution. In the next section, we analyse 2D
simulations in more detail and explore parameter space.

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}
  \delta v_R = \frac{\delta}{M}c_s(R)\sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
depend on $\phi$, and we set $M=10$.

\subsection{FARGO 2D}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$ and resolution $N_R\times N_\phi = 1024\times 2048$ or
about $16$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
approximates for the vertical disc thickness. A more appropriate
value would be $\sim H$ \citep{muller12}. However, because
$\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
(Eq. \ref{2d_grav}) is no longer symmetric. We choose a
small  
$\epsilon_g$ in favour of angular momentum conservation, keeping in
mind that the strength of self-gravity will be over-estimated.}.

\subsection{ZEUS-MP 3D}

\subsection{PLUTO 3D}
